\cursedforest\ - A Random Forest Implementation for ``Big'' and ``Wide'' Data \\ Supplementary Information

### 2017-01-18 doc/publications/plos16/plos16.tex
1000 genomes section

\marginpar{this section is not one sentence? has it been lost?}

>>>>

In a previous paper, we clustered individuals from the 1000 Genomes Project using k-means clustering\cite{Obrien2016}.
Although we were able to cluster individuals on all 80 million variants, it was relatively slow, with the entire
process (pre-processing and clustering) taking over 30 hours.
We investigate building random forest models on this data---however we observed this process to not only be slower
than building a k-means model, but also unable to scale to as many dimensions (see Table~\ref{timingtable}). To keep
the following comparisons fair, we limit the number of executors available to Spark. On the relatively low-dimensional
phase1\_chr22 dataset (1,092 x 490,036), we observed a runtime of under 8 minutes. Although slower than Ranger---
presumably due to the overhead of Spark's distributed platform on this relatively small dataset---we observe Spark ML to
be about 16 minutes faster than R. Moving to the larger dataset, phase3\_chr1 (2,504 x 6,450,364), we observe an increase
in Spark ML's runtime to over 8 hours. Moving to our next largest dataset, phase3\_chr1\-3 (2,504 x 19,328,051), and the
job fails to complete within our arbitrary cutoff time of 20 hours.

From our trials, it was apparent that Spark ML---although designed to handle many samples---is unable to efficiently cope
with a huge number of features. With the decreasing costs of genome sequencing, this case will surely become increasingly
common.

\cursedforest solves this problem of high-dimensionality genomic data. Not only is it faster on the above smaller datasets,
but it is also able to scale to the entire genome. In table~\ref{cursedforesttable} we present our observed times building
models with CursedForest using all available resources on our Spark cluster.

As we previously mentioned, building a random forest model on the phase3\_chr1 dataset with Spark ML takes over 8 hours.
Building a model with \cursedforest, however, takes only 20 minutes using the same cluster setup and this time drops to
nearly 4 minutes using the entire compute cluster. Subsequently, phase3\_chr1\-3 takes about 11 minutes, and building a
model on the entire genome is just shy of 37 minutes. Not only does the job complete successfully, but building a random
forest model on all 22 autosomes (phase3\_chr1\-22) using \cursedforest is more than twice as fast as building a random
forest on the chromosome 1 (phase3\_chr1) using Spark ML.
